
#  Tutorial on data processing and msqrob2 analysis of experiments with simple designs {-}


The result of a quantitative analysis is a list of precursor, peptide and/or protein abundances for every protein in different samples. In this tutorial we introduce a generic workflow for differential analysis of quantitative datasets with simple experimental designs.

In order to extract relevant information from these high troughput datasets, we will use our [msqrob2](https://www.bioconductor.org/packages/release/bioc/html/msqrob2.html) software tool [@goeminne2016], [@goeminne2020] and [@sticker2020]. 

# Staes spike-in study (DIA-NN output)

We will use using a publicly available spike-in study published by Staes et al. [@Staes2024]. 
They spiked digested UPS proteins in a yeast digested background at the following ratio's (yeast:ups ratio 10:1, 10:2, 10:4, 10:8, 10:10).
Here we will use a subset of the data, i.e. dilutions 10:2 and 10:4. 

We will use output of the search engine DIA-NN 2.2.0. 
The main search output for this DIA-NN version was stored in the report.parquet file in the DIA-NN output directory, which can be found under data/spikein24-staesetal2024.parquet

DIA-NN provides multiple quantifications, e.g. derived from the MS1 or MS2 spectra, and at precursor or protein (protein group) level. The term 'precursor' refers to a charged peptide species and is the basic unit of identification and quantification in DIA. Hence, in the context of DIA we refer to a precursor table, instead of to a PSM table in DDA. 

Examples of different quantities are:

- raw MS1 area: Ms1.Area, normalised MS1 Area: Ms1.Normalised, MS2 Precursor quantities: Precursor.Quantity, Normalised MS2 Precursor quantities: Precursor.Normalised, etc., which are all at the precursor level 
- MS2 based summary at the protein (protein group)-level: PG.MaxLFQ



[1.a] Participants can perform an analysis using the [Rmarkdown script](./staes-median-maxLFQ.html). Here, we will use the `Precursor.Quantity` column. Follow the steps in the script and try to understand each of the analysis steps. We know the real FC for the spike in proteins and the yeast proteins (see description of the data). What do you observe?

[1.b] Repeat the analysis and change the normalisation by using normalisation factors calculated based on the total intensity, i.e. by using the `nf_log_tot` function with arguments `nf_log_tot(qf, "precursors_log")`. What do you observe and try to explain this.

[1.c] Repeat the analysis starting from the `Precursor.Normalised` column. What do you observe? Are all data processing steps needed?

[1.d] Repeat the analysis, again use the basis script wiht median normalisation factors (`nf_log_med` function). First change the summarisation with the `aggregateFeatures` function by replacing maxLFQ with median polish (`fun = MsCoreUtils::medianPolish`), then use simple median summarisation (`fun = matrixStats::colMedians`).  (Note, that you also have to add an additional argument for median polish and median summarisation to handle missing values, i.e. add an additional argument `na.rm = TRUE`, i.e. replace `fun = function(X) iq::maxLFQ(X)$estimate` with fun = MsCoreUtils::medianPolish, na.rm = TRUE`). What do you observe and try to explain this. 

# Breast cancer example

Eighteen Estrogen Receptor Positive Breast cancer tissues from from patients treated with tamoxifen upon recurrence have been assessed in a proteomics study. Nine patients had a good outcome (or) and the other nine had a poor outcome (pd).
The proteomes have been assessed using an LTQ-Orbitrap in data dependent acquisition (DDA) mode and the thermo output .RAW files were searched with MaxQuant (version 1.4.1.2) against the human proteome database (FASTA version 2012-09, human canonical proteome). 

The data are already processed and stored as QFeatures objects. They can be loaded in R using readRDS so you can immediately start at the data modeling. So you only have to read the data and you can start with section 5 Data exploration in the basic script. 

```{r eval=FALSE}
qf <- readRDS(url("https://github.com/statOmics/PDA-DIA/raw/refs/heads/main/data/cancer3x3.rds","rb"))
```

Three QFeatures objects are available (you can simply modify the url above replacing the filename):

1. For a 3 vs 3 comparison: cancer3x3.rds
2. For a 6 vs 6 comparison: cancer6x6.rds
3. For a 9 vs 9 comparison: cancer9x9.rds

# Staes spike-in study - Spectronaut output

Redo the data analysis for the spectronaut output file. 
Modify [Rmarkdown script](./staes-median-maxLFQ.html). 
Keep the same processing steps. 
Below you can see the mappings of the DIA-NN variables to their spectronaut counterparts.

```{r eval = FALSE}
precursors <- precursors |> 
  select(
         R.FileName, #Run, 
         EG.PrecursorId, #Precursor.Id, 
         EG.ModifiedSequence, #Modified.Sequence, 
         PEP.StrippedSequence, #Stripped.Sequence, 
         FG.Charge, #Precursor.Charge, 
         PG.ProteinGroups, #Protein.Group, 
         PG.ProteinNames, #Protein.Names,
         PG.ProteinAccessions, #Protein.Ids,
         PG.Genes, #Genes,
         FG.MS2RawQuantity, #Precursor.Quantity,
         FG.MS2Quantity, #No DIA-NN counterpart 
         FG.Quantity, #Precursor.Normalised, 
         FG.MS1RawQuantity, #Ms1.Area,
         FG.MS1Quantity, #Ms1.Normalised #No real DIA-NN counterpart 
         EG.Qvalue, #Q.Value, 
         #No spectronaut counterpart #Lib.Q.Value,
         PG.Qvalue, # PG.Q.Value,
         #No spectronaut counterpart #Lib.PG.Q.Value
         PEP.IsProteotypic, #Proteotypic,
         EG.IsDecoy, #Decoy,
         EG.ApexRT) #RT)
```


### References